On the Minimum Core Mass for Giant Planet Formation at Wide Separations

### 2014-02-13 tex/PaperI/ms.tex
more edits, final significant ones hopefully

This work develops quasi-static models in which the core has a fixed mass and is no longer accreting solids.  By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.  Previous studies have considered this type of growth \citep{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.  Especially since phase 2 atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is of crucial importance.  Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.

@@

\textbf{A larger disk lifetime further reduces $M_{\rm crit}$. Recent studies have shown that gas disks may live up to $\propto$$10-12$ Myr \citep{bell13}. As $M_{\rm crit} \sim t\di^{-3/5}$ (cf. Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.  Of course we also require some fraction of the disk lifetime for the core to grow and possibly migrate.}

@@

\textbf{Our results are consistent with those of previous core accretion studies. The solution of \citet{stevenson82}, which considers fully radiative atmospheres with constant opacity, predicts that the critical core mass is lower for lower envelope opacity and higher mean molecular weight of the nebular gas. Additionally, \citet{stevenson82} finds that the critical core mass decreases weakly with increasing envelope radius (via a logarithmic factor), i.e. with increasing orbital distance. Both of the above qualitative predictions are reproduced by our result.}

\textbf{In the regime of no ongoing planetesimal accretion, \citet{ikoma00} find an approximate analytic expression for the envelope formation time (the equivalent of the runaway accretion time $t_{\rm run}$ in our models), }

\begin{equation}
\label{eq:tauenv}
\tau_{\rm env} \sim 3 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.5} \Big(\frac{\kappa}{1 \text{cm$^2$ s$^{-1}$}}\Big) \,\,\, \text{yr},
\end{equation}
\textbf{at $a=5.2$ AU, Jupiter's location. The result above reproduces the linear opacity dependence of the atmospheric evolutionary times in our model (see \S\ref{sec:timeev}).  Moreover, we can approximate $t_{\rm run}$ in our model at 5 AU as} 
\begin{equation}
\label{eq:trun5au}
t_{\rm run} \sim 5 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.4} \,\,\, \text{yr},
\end{equation} 
\textbf{for the opacity-law (\ref{eq:opacitylaw}). Our results are thus consistent with those of \citet{ikoma00} at 5 AU, both in terms of scaling and magnitude.}

\textbf{Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work. \citet{pollack96} find an evolutionary time and crossover mass of $\sim$$7.5$ Myrs and $\sim$$16 M_{\oplus}$, respectively, for a MMSN disk model at 5.2 AU and interstellar grain opacity. These values, however, decrease to $\sim$$3$ Myrs and $\sim$$12 M_{\oplus}$ if planetesimal accretion is entirely shut off during the gas accretion phase. The core accretion model of \citet{HubBod05} for Jupiter's formation predicts an evolutionary time of 3.3 Myrs for a $10 M_{\oplus}$ core, for an interstellar opacity, which is also consistent with our result at 5 AU.}

%\textbf{Lastly, the decline in $M_{\rm crit}$ with semimajor axis is also found by R06 for planets forming at $\gtrsim 5$ AU. A quantitative comparison with our results requires detailed calculations of the }

@@

Rapid gas accretion onto low mass cores could explain the origin of distant directly imaged giant planets \citep{marois08, lagrange10}.  However, our model does not address the details of how solid cores grow, and many possibilities exist. \textbf{Our brief treatment of core formation is purposeful, as there are large uncertainties in both planetesimal formation and the subsequent processes of core growth, particularly at large orbital distances. Core formation in the outer disk is challenging due to large dynamical times, and cores are unlikely to reach critical masses at $\gtrsim 40-50$ AU under classic core accretion scenarios \citep{Raf11}. However, modern models of aerodynamic pebble accretion (\citealt{ormel10}; \citealt{lambrechts12}) can give faster growth. Moreover, cores can form closer to the star, then migrate or be scattered outwards by other giants in the system \citep{ida13}.}

 For a giant planet to form with a core near the minimum mass, the core's growth must  first be rapid and then slow significantly, as in phases I and II, respectively, of \cite{pollack96}. Initial core growth must be fast, compared to the disk lifetime, to get a sufficiently massive core.  Subsequent slowing of core growth is required to allow the atmosphere to cool and contract.  To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.  Specifically, a core mass doubling timescale of $\sim$$400$ Myr, or faster, would provide enough heating to stall atmospheric cooling and growth.

Thus for giant planets to form around cores near $\MC$, core growth must stall quite severely.   The accretion and scattering of local planetesimals, combined with the inward aerodynamic drift of smaller solids past the core, offers a plausible means to sufficiently isolate the core.

On the other hand, non-negligible accretion of solids onto cores could continue for the lifetime of the gas disk. \textbf{For our fiducial disk model, the isolation mass, $M_{\rm iso} \propto a^{3/4}$, increases with orbital distance, and may thus become larger than $M_{\rm crit}$ at large separations. In this case, the ongoing accretion of solids increases the critical core mass by inhibiting atmospheric growth, and may prevent accumulation of a significant envelope before the disk dissipates. It follows that} relatively high mass cores would remain as ice giants, akin to Uranus and Neptune. \textbf{Non-traditional accretion models such as pebble accretion or scattering are thus needed in the outer disk in order to avoid slow growth to the isolation mass, which would stall atmospheric evolution.}  Ongoing direct imaging surveys and their successors \citep{hinz12, macintosh12} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.

\acknowledgements{We thank Ruth Murray-Clay for detailed feedback and skillful proof reading. We appreciate valuable comments from Eugene Chiang and Scott Kenyon.  ANY thanks Phil Armitage for stimulating conversations.

>>>>

This work develops quasi-static models in which the core has a fixed mass and is no longer accreting solids.  By isolating the process of KH contraction, we determine the minimum timescale for runaway atmospheric growth, and thus for giant planet formation by core accretion.  Previous studies have considered this type of growth \citep[hereafter I00; PN05, respectively]{ikoma00, pn05}, which is an end-member case of phases 2 and 3 in more detailed quasi-static models.  Especially since phase 2 atmospheric growth is often the slowest stage of core accretion, understanding the fastest allowed rates of gas accretion is of crucial importance.  Our study develops simplified cooling models for the purposes of broadly exploring parameter space -- especially in the outer disk -- and developing a more detailed understanding of atmospheric accretion.

@@

\textbf{A larger disk lifetime further reduces $M_{\rm crit}$. Recent studies have shown that gas disks may live up to $\propto$$10-12$ Myr \citep{bell13}. As $M_{\rm crit} \sim t\di^{-3/5}$ (cf. Equation \ref{eq:tcoolan}), a disk lifetime of 10 Myrs decreases the critical core mass by a factor of two.  Of course, some fraction of the disk lifetime must be allocated to the growth and possibly migration of the core.}

@@

We can directly compare our results with other studies of protoplanetary atmospheric growth in the absence of planetesimal accretion, notably those in I00 and PN05.  A major distinguishing feature of our study is that we study a range of disk conditions, while I00 and PN05 focused on growth at the location of Jupiter, 5.2 AU, in their disk models.  Also both I00 and PN05 solved the full time dependent energy equation and considered more detailed EOSs, in addition to other detailed differences in model parameters.  Despite these differences, the qualitative and quantitative agreement with our study is good.

\textbf{I00 give an approximate analytic fit to their models' envelope formation time (the equivalent of the runaway accretion time $t_{\rm run}$ in our models), at $a=5.2$ AU, }
\begin{equation}
\label{eq:tauenv}
\tau_{\rm env} \sim 3 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.5} \Big(\frac{\kappa}{1 \text{cm$^2$ s$^{-1}$}}\Big) \,\,\, \text{yr}\, .
\end{equation}
\textbf{Our model reproduces the linear opacity dependence (see \S\ref{sec:timeev}), even though we use a temperature dependent opacity, \Eq{eq:opacitylaw}, instead of a constant $\kappa$.  Our growth times at 5 AU,} 
\begin{equation}
\label{eq:trun5au}
t_{\rm run} \sim 5 \times 10^8 \Big(\frac{M\co}{M_{\oplus}}\Big)^{-2.4} \,\,\, \text{yr},
\end{equation} 
\textbf{agree well with the I00 results, both in magnitude and scaling with core mass.}

One goal of our study was to explain the extended luminosity minimum in evolutionary models that characterizes phase 2 in core accretion models as described in the introduction.  Previous work (see Figure 2 in I00 and Figures 3 and 4 in PN05) show that this luminosity minimum is an intrinsic feature of atmospheric cooling even in the absence of planetesimal accretion.  We not only reproduce this effect in a simple model (see \Figs{fig:Ltplot}{fig:LtvsMopacity}) but show that self-gravity is the essential ingredient to produce a broad luminosity minimum well before the crossover mass.

\textbf{Time-dependent studies that incorporate planetesimal accretion generally find larger formation timescales and critical core masses than those of this work.  This result is expected since additional energy from planetesimal accretion limits the ability of the atmosphere to cool. \citet{pollack96} find an evolutionary time and crossover mass of $\sim$$7.5$ Myrs and $\sim$$16 M_{\oplus}$, respectively, for a MMSN disk model at 5.2 AU and interstellar grain opacity. These values, however, decrease to $\sim$$3$ Myrs and $\sim$$12 M_{\oplus}$ if planetesimal accretion is entirely shut off during the gas accretion phase. The core accretion model of \citet{HubBod05} for Jupiter's formation predicts an evolutionary time of 3.3 Myrs for a $10 M_{\oplus}$ core, for an interstellar opacity, which is also consistent with our result at 5 AU.}

\textbf{A direct comparison with studies that only include planetesimal accretion, and neglect atmospheric cooling, is difficult.  Note however that I00 establishes the correspondence between the minimum luminosity in evolutionary models and the minimum planetesimal accretion rate needed in static models, which gives rise to the classical critical core mass of static models \citep{mizuno78, stevenson82}.}

\textbf{In terms of modern static studies, our results complement (and borrow some tools from) R06 and \citet{Raf11}.  These studies considered the disk radius dependence of core accretion for protoplanets that continuously accrete planetesimals.  We show that the limits on core accretion at large radial distance which \cite{Raf11} (and references therein) claim can be overcome if planetesimal accretion shuts off and the atmosphere is allowed to cool.  The plausibility of rapid core growth followed by negligible subsequent solid accretion admittedly remains highly uncertain, as discussed in more detail in the Summary.}

@@

Rapid gas accretion onto low mass cores could explain the origin of distant directly imaged giant planets \citep{marois08, lagrange10}.  However, our model does not address the details of how solid cores grow, as many possibilities exist and many uncertainties remain. %\textbf{Our brief treatment of core formation is purposeful, as there are large uncertainties in both planetesimal formation and the subsequent processes of core growth, particularly at large orbital distances. Core formation in the outer disk is challenging due to large dynamical times, and cores are unlikely to reach critical masses at $\gtrsim 40-50$ AU under classic core accretion scenarios \citep{Raf11}. However, modern models of aerodynamic pebble accretion (\citealt{ormel10}; \citealt{lambrechts12}) can give faster growth. Moreover, cores can form closer to the star, then migrate or be scattered outwards by other giants in the system \citep{ida13}.} 
For a giant planet to form with a core near the minimum masses we derive, core growth must  first be rapid and then slow significantly, as in phases 1 and 2, respectively, of \cite{pollack96}.

Initial core growth must be fast, compared to the disk lifetime, to get a sufficiently massive core.  \textbf{Such rapid core growth is possible in a variety of scenarios, including the fastest gas-free planetesimal accretion rates \citep{dones93} and -- probably more relevantly for gas rich disks --  the aerodynamic accretion of mm-m sized ``pebbles" and ``boulders" in gas disks \citep{ormel10, lambrechts12}.  Moreover, cores could form rapidly closer to the star, then migrate or be scattered outwards by already formed giants \citep{ida13}.}

\textbf{A stronger constraint is that} core growth subsequently slow severely, to allow the atmosphere to cool and contract.  To be more quantitative, the minimum cooling luminosity in \Fig{fig:Ltplot}, $L\approx 3.5 \times 10^{24}\;\mathrm{erg s}^{-1}$, could be cancelled by low levels of heating from solid accretion.   A core mass doubling timescale of $\sim$$400$ Myr, or faster, would thus provide enough heating to stall atmospheric cooling and growth.   \textbf{An additional concern is that isolation masses tend to grow with disk radius, as $M_{\rm iso} \propto \varSigma_{\rm p}^{3/2} a^2 \propto a^{3/4}$ under the approximation that the surface density of accreted planetesimals ($\varSigma_{\rm p}$) scales with the gas \citep{youdin13}.   While this behavior is nominally inconsistent with final core masses that decline with distance, the predictive power of the isolation mass is imperfect.   For starters, the efficiency of planetesimal formation remains uncertain.  Moreover the locality of core growth, which underlies the isolation mass, disappears when accreted solids drift and/or cores migrate significantly. }

Thus while our calculations show that low mass cores can grow into gas giants in the outer disk, ongoing solid accretion could prevent significant atmospheric growth.  In the Solar System, the ice giants Uranus and Neptune,  with core (here ice and rock) masses of $\sim 13 -15$ M$_\oplus$, argue for the latter possibility.  Ongoing exoplanet imaging surveys and their successors \citep{hinz12, macintosh12, close14} will help discriminate among the various planet formation pathways in the outskirts of protoplanetary disks.

\vspace*{5 mm}

\acknowledgements{We thank the referee for a thorough report and Ruth Murray-Clay for detailed feedback and skillful proof reading. We appreciate valuable comments from Eugene Chiang and Scott Kenyon.  ANY thanks Phil Armitage for stimulating conversations.
